# Bootstrapping and Confidence Intervals  {#ch13}

## Learning objectives

By the end of this chapter, you should be able to:

- understand why sample percentages differ from population percentages
- calculate and interpret the standard error of a percentage
- construct confidence intervals using the empirical rule
- correctly interpret what a confidence interval does and does not mean
- distinguish between point estimates and interval estimates

## Motivation: Should the politician run?

A politician is deciding whether to enter an election. In a village of 100,000 eligible voters, a survey of 2,500 people finds that 1,328 support the candidate, which is about 53%.

Should the politician run?

At first glance, the answer seems obvious: 53% is greater than 50%. But this conclusion depends on how accurate the estimate is. The key question is:

> How far might the sample percentage be from the true population percentage?

::: {.callout-note title="Key idea"}
A sample percentage is only an estimate. It is subject to **chance error** due to random sampling.
:::

## From samples to uncertainty

The sample percentage, 53%, is a **point estimate** of the population percentage. But different random samples would give slightly different results.

To quantify this uncertainty, we use the **standard error (SE)**.

For a percentage, the standard error is:

$$
SE(\%) = \frac{SD(\text{box})}{\sqrt{n}} \times 100
$$

But there is a problem: we do not know the standard deviation of the box/population.

### Estimating variability from the sample

We go around this problem by using the sample itself to estimate variability.

In this case:

- 53% of observations are 1 (support)
- 47% are 0 (do not support)

The standard deviation of the sample, which we denote $SD^+$ of such a 0--1 variable is approximately:

$$
SD^+ = (1-0) \sqrt{0.53 \times 0.47} \approx 0.5
$$

Using this, the standard error becomes approximately:

$$
SE(\%) = \frac{SD^+}{n} \times 100\% \approx 1\%
$$

So the estimate of 53% is likely to be off by about 1%.

::: {.callout-tip title="Interpretation"}
Being off by 3% would correspond to about 3 standard errors, which is unlikely.
:::

**Conclusion:** The candidate is very likely above 50% support and may reasonably decide to run.

## Confidence intervals

Sometimes instead of a **point estimate**, we can provide an **interval estiate** to account for uncertainty.

A **confidence interval** provides a range of plausible values for the population percentage.

Using the empirical rule:

::: {.callout-note title="Confidence Interval"}
- $\pm 1$ SE gives about 68% confidence
- $\pm 2$ SE gives about 95% confidence
- $\pm 3$ SE gives about 99.7% confidence
:::

In our example:

$$
53\% \pm 2\% = [51\%, 55\%]
$$

That is, a 95% confidence interval is approximately:

$$
\text{estimate} \pm 2 \times SE
$$


### A common misunderstanding

It is tempting to say:

> There is a 95% chance that the true percentage lies between 51% and 55%.

This is incorrect.

::: {.callout-warning title="Common pitfall"}
The population percentage is fixed. It does not vary.

The interval varies because it depends on the sample.
:::

The correct interpretation is:

> If we repeatedly took samples and constructed confidence intervals, about 95% of those intervals would contain the true population percentage.

### Confidence intervals as repeated sampling

Imagine drawing 100 different samples of 2,500 voters.

Each sample gives:

- a different percentage
- a different confidence interval

Some intervals will include the true value. Others will miss it.

::: {.callout-note title="Key idea"}
A 95% confidence level means that about 95 out of 100 such intervals will capture the true parameter.
:::

### Simulation: One poll

Let us simulate one poll where the true population support rate is 80%, and each sample contains 2,500 observations.

```{r}
set.seed(123)

n <- 2500
p_true <- 0.80

sample_draw <- rbinom(n, size = 1, prob = p_true)
p_hat <- mean(sample_draw)
se_hat <- sqrt(p_hat * (1 - p_hat) / n)

p_hat
se_hat
```

The estimated percentage is:

```{r}
round(100 * p_hat, 2)
```

The estimated standard error in percentage points is:

```{r}
round(100 * se_hat, 2)
```

The approximate 95% confidence interval is:

```{r}
lower <- p_hat - 2 * se_hat
upper <- p_hat + 2 * se_hat

round(100 * c(lower, upper), 2)
```

### Simulation: 100 confidence intervals

Now let us simulate 100 independent samples, each of size 2,500, from a population where the true percentage is 80%.

```{r}
#| echo: false
set.seed(123)

n_sims <- 100
n <- 2500
p_true <- 0.80

results <- data.frame(
  sample_id = 1:n_sims,
  p_hat = NA_real_,
  se_hat = NA_real_,
  lower = NA_real_,
  upper = NA_real_,
  covers = NA
)

for (i in 1:n_sims) {
  x <- rbinom(n, size = 1, prob = p_true)
  p_hat_i <- mean(x)
  se_hat_i <- sqrt(p_hat_i * (1 - p_hat_i) / n)
  lower_i <- p_hat_i - 2 * se_hat_i
  upper_i <- p_hat_i + 2 * se_hat_i
  covers_i <- (lower_i <= p_true) & (upper_i >= p_true)
  
  results$p_hat[i] <- p_hat_i
  results$se_hat[i] <- se_hat_i
  results$lower[i] <- lower_i
  results$upper[i] <- upper_i
  results$covers[i] <- covers_i
}

head(results)
```

How many of these 100 intervals cover the true population percentage?

```{r}
sum(results$covers)
```

What percentage of intervals cover the true value?

```{r}
mean(results$covers) * 100
```

### Plot: 100 confidence intervals

```{r}
#| echo: false
plot(
  x = NULL, y = NULL,
  xlim = c(min(results$lower), max(results$upper)),
  ylim = c(1, n_sims),
  xlab = "Estimated population percentage",
  ylab = "Sample number",
  main = "95% confidence intervals for 100 samples",
  xaxt = "n"
)

axis(1, at = seq(0.74, 0.86, by = 0.02), labels = paste0(seq(74, 86, by = 2), "%"))

for (i in 1:n_sims) {
  if (results$covers[i]) {
    segments(results$lower[i], i, results$upper[i], i)
    points(results$p_hat[i], i, pch = 16)
  } else {
    segments(results$lower[i], i, results$upper[i], i, lty = 2, lwd = 2)
    points(results$p_hat[i], i, pch = 1, cex = 1.1)
  }
}

abline(v = p_true, lwd = 2)
```

::: {.callout-tip title="How to read the figure"}
Each horizontal line is a 95% confidence interval from one sample.

* The vertical line marks the true population percentage, 80%.
* Intervals that cross the vertical line capture the true value.
* Intervals that do not cross it miss the true value.
:::

### Why some intervals fail

Even when the method is correct, some intervals will fail to include the true value. That is not a mistake. It is exactly what a 95% confidence level implies: about 5% of intervals will miss the true value just by chance.

::: {.callout-important title="Important"}
A confidence procedure can be reliable without guaranteeing success in every single sample.
:::

### Deduction vs induction

There are two directions of reasoning in statistics:

* **Deduction:** from population to sample
* **Induction:** from sample back to population

Confidence intervals are a tool for **inductive reasoning**.

::: {.callout-note title="Key idea"}
Statistical inference is about learning about the population from a sample.
:::

## Chapter summary

* Sample percentages vary due to chance error.
* The standard error measures the typical size of this variation.
* Confidence intervals provide a range of plausible values for the population parameter.
* A 95% confidence interval captures the true value in about 95% of repeated samples.
* The interval varies across samples; the parameter does not.

## Exercises

### 1. Graduate earnings {.unnumbered}

A simple random sample of 100 graduates from a certain college found that 48 were earning **Baht 50,000 per month or more**.

1. Estimate the percentage of all graduates from that college who earn Baht 50,000 per month or more.
2. Attach a “give-or-take” number to your estimate by computing the standard error of the percentage.
3. Write one sentence interpreting your result in plain language.

---

### 2. Standard error from a box model {.unnumbered}

Suppose there is a box of 100,000 tickets, each marked either 0 or 1. In fact, **20%** of the tickets are 1s.

A sample of **400 draws** is taken at random.

1. What is the expected percentage of 1s in the sample?
2. Calculate the standard error of the percentage of 1s.
3. Explain briefly why the standard error becomes smaller when the sample size increases.

---

### 3. Internet use among subscribers {.unnumbered}

TRUE company serves **1,000,000 subscribers**. As part of a market survey, a simple random sample of **3,600 people** was taken. In the sample:

- 1,080 use the internet for banking
- 720 use the internet for reading the news
- 1,440 use the internet for online games

Answer the following:

1. What percentage of customers use the internet for:
   - banking,
   - reading the news,
   - online games?
2. Compute the standard error for each percentage.
3. For internet banking, what is the approximate probability that the sample percentage lies between **28% and 32%**?

---

### 4. Polling support for a political party {.unnumbered}

In a poll of **200 voters**, 92 said that they would vote for the **YY Party**.

1. Estimate the percentage of all voters who support the YY Party.
2. Construct an approximate **95% confidence interval** for the true population percentage.
3. Do you think this estimate is accurate enough to advise the YY Party to enter the election?
4. How could the accuracy of the estimate be improved?

---

### 5. Percentages in a town population {.unnumbered}

According to a census, a certain town has **100,000 people aged 18 and over**. Of these:

- **60%** are married
- **10%** earn more than **Baht 75,000 per month**
- **20%** have university degrees

A simple random sample of **1,600 people** will be drawn from this population.

#### (a) Marriage {.unnumbered}

To study marriage, a box model is needed.

1. Should the box contain **1,600 tickets** or **100,000 tickets**? Explain.
2. What is the chance that **58%** of the sample are married?

#### (b) High income {.unnumbered}

To study the proportion earning more than Baht 75,000 per month:

1. Should each ticket in the box show the person’s exact income? Explain.
2. What is the chance that **11% or more** of the sample earn above Baht 75,000 per month?

#### (c) University degree {.unnumbered}

Find the chance that **between 19% and 21%** of the sample have a university degree.

---

### 6. Health insurance {.unnumbered}

A survey of **400 people** finds that **80** have health insurance.

1. Estimate the percentage of the population with health insurance.
2. Construct a **90% confidence interval** for the population percentage.
3. Explain what your interval means.

---

### 7. Consumer survey: televisions and VCRs {.unnumbered}

A utility company serves **50,000 households**. As part of a survey of consumer attitudes, it takes a simple random sample of **750 households**. The average number of television sets in the sample is **1.86**, and the sample SD is **0.80**.

#### (a) Average number of television sets {.unnumbered}

1. If possible, construct a **95% confidence interval** for the average number of television sets in all 50,000 households.
2. If this is not possible, explain why.

#### (b) Percentage with VCRs {.unnumbered}

Out of the 750 households in the survey, **351 have VCRs**.

1. Estimate the percentage of all 50,000 households with VCRs.
2. If possible, construct a **99.7% confidence interval** for that percentage.

#### (c) Households without a television {.unnumbered}

Among those surveyed, **749 households have at least one television set**.

1. Estimate the percentage of households with **no television set**.
2. If possible, construct a **68% confidence interval** for the percentage of all 50,000 households with no television set.
3. If this is not possible, explain why.

---

### 8. Red and blue marbles {.unnumbered}

A box contains **1 red marble** and **99 blue marbles**. Ten marbles are drawn at random **with replacement**.

1. Find the **expected number** of red marbles in the 10 draws.
2. Find the **standard error** of the number of red marbles.
3. What is the chance of drawing **fewer than 0 red marbles**?
4. Use the normal curve to estimate this chance.
5. Does the probability histogram for the number of red marbles look like a normal curve? Explain.

---

### Conceptual question

### 9. Point estimates and confidence intervals {.unnumbered}

For each statement below, say whether it is **true or false**, and explain briefly.

1. A sample percentage is usually equal to the population percentage.
2. A larger sample gives a smaller standard error.
3. A 95% confidence interval means there is a 95% chance that the true value is inside the interval.
4. If we repeatedly draw samples and construct 95% confidence intervals, about 95% of them will contain the true population value.

**(a)** Why do statisticians usually prefer a **confidence interval** to a **point estimate** when reporting survey results?

**(b)** Explain in your own words: Why is it incorrect to say “there is a 95% chance the true value lies in the interval?

<!--

## Optional extension: Using ggplot2

If you prefer a cleaner plot, here is a `ggplot2` version.

```{r}
library(ggplot2)

ggplot(results, aes(y = sample_id, x = p_hat)) +
  geom_segment(aes(x = lower, xend = upper, yend = sample_id, linetype = covers)) +
  geom_point(aes(shape = covers)) +
  geom_vline(xintercept = p_true) +
  scale_x_continuous(labels = function(x) paste0(round(100 * x), "%")) +
  labs(
    title = "95% confidence intervals for 100 samples",
    x = "Estimated population percentage",
    y = "Sample number"
  ) +
  theme_minimal()
```

-->

